---
title: 'Are Roger and Serena the all-time best tennis players?'
author: 'Your Teacher Today: Bobby Stuijfzand'
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Before we begin

### Some steps to follow if you want to code along

1. Go to [https://github.com/BobbyGlennS/fds-learner-lab](https://github.com/BobbyGlennS/fds-learner-lab) and download the repository.
1. Unzip the repository and open the `fds-learner-lab.Rproj` file. RStudio should now start.
2. From the `files` viewer inside RStudio, open the `leaner.Rmd` file.

### Beginning our R code

Load packages

```{r}
library(lubridate)
library(tidyverse)
```

## Defining the question

In this webinar we will use data from the WTP and ATP to answer the question:

***Who are the best tennis players of all-time?***

### The scope

We will limit this question to each gender separately - we are not going to compare between genders.

We will further look only at who won the grand slam single tournaments.

### Sub questions

We will use two questions to get some insight into our main question.

- Who won the most grand slam tournaments (single player)?
- Who won the most grand slam tournaments consecutively?

After we have found answer to these questions, we can combine this information to give a comprehensive answer to our main question.

## The data

### Credit where credit is due

The dataset we will be using is carefully assembled by [Jeff Sackmann](https://github.com/JeffSackmann) and is freely available on the internet in the form of github repositories.
For reference - you can find the sources for the full women and men data respectively here:

- [https://github.com/JeffSackmann/tennis_wta](https://github.com/JeffSackmann/tennis_wta)
- [https://github.com/JeffSackmann/tennis_atp](https://github.com/JeffSackmann/tennis_atp)

For this webinar I have already done all the pre-processing so we can cut straight to the chase.

### Load the data

The data can be found in the `data` folder in the repository.
There is a file for the women (`wta_grandslam.csv`) and one for the men (`atp_grandslam.csv`).

We'll load this here.

```{r}
data_wta <- read_csv("data/wta_grandslam.csv")
data_atp <- read_csv("data/atp_grandslam.csv")
```

Let's now get a first glimpse of the data: 

```{r}
head(data_wta) %>% knitr::kable()
```

### Adding dates

Change the `tourney_date` column into a date type and sort our data.

```{r}
data_wta <- data_wta %>% 
  mutate(date = ymd(tourney_date)) %>% 
  arrange(date)

data_atp <- data_atp %>% 
  mutate(date = ymd(tourney_date)) %>% 
  arrange(date)
```

Our data is now ready.

## First question: Who won the most grand slam tournaments (single player)?

---

#### EXERCISE

What code would you use to get the top 5 winners for each gender?

```{r}
# women
top5_wta <- data_wta %>%
  # add your code here...
```

```{r}
# men
top5_atp <- data_atp %>%
  # add your code here...

```

### Add a visualisation

Let's visualise these results.

```{r}
bind_rows(
  "women" = top5_wta,
  "men" = top5_atp,
  .id = "gender"
) %>% 
  mutate(winner_name = fct_inorder(winner_name),
         winner_name = fct_rev(winner_name)) %>% 
  ggplot(aes(x = n, y = winner_name)) +
  geom_col(fill = "#C0C663") +
  facet_wrap(vars(gender), scale = "free") +
  hrbrthemes::theme_ipsum_rc() +
  labs(title = "Top 5 Grand Slam Winners by Gender",
       x = "Number of titles won",
       y = "")
```

---

### Does it change with the open era?

```{r}
# women
top5_wta_open_era <- data_wta %>%
  # add your code here...
```

```{r}
# men
top5_atp_open_era <- data_atp %>%
  # add your code here...
```

### Add a visualisation

Let's visualise these results.

```{r}
bind_rows(
  "women" = top_wta_open_era,
  "men" = top_atp_open_era,
  .id = "gender"
) %>% 
  mutate(winner_name = fct_inorder(winner_name),
         winner_name = fct_rev(winner_name)) %>% 
  ggplot(aes(x = n, y = winner_name)) +
  geom_col(fill = "#C0C663") +
  facet_wrap(vars(gender), scale = "free") +
  hrbrthemes::theme_ipsum_rc() +
  labs(title = "Top 5 Grand Slam Winners by Gender",
       subtitle = "Only Grand Slam's from the Open Era are Included",
       x = "Number of titles won",
       y = "")
```


## Second question: Who won the most grand slam tournaments consecutively?

We need to keep the following things in mind:

- An athlete can win once, then lose, and then win again. We don't want these two wins to count towards the same streak.
- An athlete can have multiple streaks, so the answer will not be as unambiguous as in the previous part.

I will demonstrate an approach in steps for the women.
Then you can use this as an exercise to solve it for the men.

### Step 1: Number each game consecutively

We will use `row_number()` for this.

```{r}
data_wta_cons <- data_wta %>% 
  #code here to create `game_id`
```

We now have a number for each game:

```{r}
data_wta_cons %>% 
  head(10) %>% 
  select(tourney_name, date, game_id) %>% 
  knitr::kable()
```

### Step 2: Group our data by athlete

```{r}
data_wta_cons <- data_wta_cons %>% 
  # code here
```

Before we continue, let's have a look at the data.

```{r}
data_wta_cons %>% 
  arrange(winner_name) %>% 
  head(10) %>% 
  select(winner_name, game_id, date) %>% 
  knitr::kable()
```

### Step 3: Find the id of the previous game_id

We can use `lag()` to find the previous `game_id` an athlete has won.
We use `default = -99` so that if there is no previous game, we still have a value to work with.

```{r}
data_wta_cons <- data_wta_cons %>% 
  # code here to create variable `previous_game`
```

Let's examine the data again:

```{r}
data_wta_cons %>% 
  arrange(winner_name) %>% 
  head(10) %>% 
  select(winner_name, game_id, previous_game, date) %>% 
  knitr::kable()  
```

### Step 4: Is the current id the start of something new?

```{r}
data_wta_cons <- data_wta_cons %>% 
  # code here to create `streak_begin`
```

Let's examine the data again:

```{r}
data_wta_cons %>% 
  arrange(winner_name) %>% 
  head(10) %>% 
  select(winner_name, game_id, previous_game, streak_begin) %>% 
  knitr::kable()
```

### Intermezzo: the cumulative sum

For the next step, we are going to process the `streak_begin` variable in a way that we will get a `streak_id` for each separate streak.
In order to do so we can use a function called `cumsum()` which calculates the cumulative sum for a variable.

What is the cumulative sum?

Let's demonstrate.

We have a vector containing the values `1, 2, 3, 4`.

Calculating the cumulative sum gives:

```{r}
cumsum(c(1,2,3,4))
```

The cumulative sum returns a vector of the same length, that contains for each element the sum of the original vector *up until that point*.

How is this useful?

Consider that `TRUE` and `FALSE` are represented by `1` and `0` in R.
A cumulative sum on this vector of `0`s and `1`s:

```{r}
cumsum(c(0, 1, 0, 0, 1))
```

... will yield the same result as a cumulative sum on this *logical* vector of `FALSE`s and `TRUE`s.

```{r}
cumsum(c(FALSE, TRUE, FALSE, FALSE, TRUE))
```

You can see that only once a `TRUE` appears will the cumulative sum increment.

Now recall that in our `streak_begin` variable, we have a `TRUE` for every start of a streak.

Let's look at the data again:

```{r}
data_wta_cons %>% 
  arrange(winner_name) %>% 
  head(10) %>% 
  select(winner_name, game_id, previous_game, streak_begin) %>% 
  knitr::kable()
```

### Step 5: Get an id for each streak

If we run `cumsum()` on `streak_begin`, what we will get is a number that increments with every new beginning.

```{r}
data_wta_cons <- data_wta_cons %>% 
  # code to run cumsum() on `streak_begin` to create `streak_id`
```

Let's admire our result.

```{r}
data_wta_cons %>% 
  arrange(winner_name) %>% 
  head(10) %>% 
  select(winner_name, game_id, previous_game, streak_begin, streak_id) %>% 
  knitr::kable()
```

### Step 6: Let's count the streaks!

Now all that's left is counting up the streak ids to understand how many rows belonged to each streak, and therefore, how long each streak was!

```{r}
data_wta_cons_result <- data_wta_cons %>% 
  # code to count streaks, don't forget to sort!
```

Let's admire the result

```{r}
data_wta_cons_result %>% 
  head(20) %>% 
  knitr::kable()
```

As we stored the results in a separate tibble, We can now also go back to the original tibble to examine some individual streaks.

For example Serena Williams' longest streak:

```{r}
data_wta_cons %>% 
  filter(streak_id == 13) %>% 
  select(winner_name, date, tourney_name)
```

## And now for the men...

---

#### EXERCISE: 

Can you answer the second question for the men?

---

## Open era tennis results

---

#### BONUS:

What would the results look like if we account for the switch to allowing professionals to compete in grand slams?

Hint: the open era, as it was called, started in 1968.

---
